---
title: "script"
output: pdf_document
date: "2025-12-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploratory data analysis


```{r cars}
# Load packages
library(dplyr)
library(ggplot2)
```

Reading in the dataset:

```{r}
dat <- read.csv("student-mat.csv", sep = ";")
str(dat)

```

Organizing into relevant cols:


```{r}
dat <- dat %>%
  mutate(
    sex      = factor(sex),
    school   = factor(school),
    address  = factor(address),
    famsize  = factor(famsize),
    Pstatus  = factor(Pstatus),
    internet = factor(internet),
    romantic = factor(romantic),
    studytime = factor(
      studytime,
      levels = 1:4,
      labels = c("low", "moderate", "high", "very high")
    )
  )

# Basic summary of numeric variables
summary(select(dat, G1, G2, G3, age, absences))

```

Distribution of math grades?


```{r}

ggplot(dat, aes(x = G3)) +
  geom_histogram(binwidth = 1, boundary = 0, closed = "left") +
  labs(
    title = "Distribution of Final Math Grades (G3)",
    x = "Final grade (0–20)",
    y = "Number of students"
  )

```

Grades by gender

```{r}

ggplot(dat, aes(x = sex, y = G3)) +
  geom_boxplot() +
  labs(
    title = "Final Grade (G3) by Sex",
    x = "Sex",
    y = "Final grade (G3)"
  )

```


Grades by study time:


```{r}

ggplot(dat, aes(x = studytime, y = G3)) +
  geom_boxplot() +
  labs(
    title = "Final Grade (G3) by Study Time Category",
    x = "Weekly study time",
    y = "Final grade (G3)"
  )

```

Relationship between grades at different time points


```{r}

#g1
ggplot(dat, aes(x = G1, y = G3)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "G1 vs G3",
    x = "First period grade (G1)",
    y = "Final grade (G3)"
  )


#g2
ggplot(dat, aes(x = G2, y = G3)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "G2 vs G3",
    x = "Second period grade (G2)",
    y = "Final grade (G3)"
  )

```


Relationship between absences and final grade:


```{r}

ggplot(dat, aes(x = absences, y = G3)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Absences vs Final Grade",
    x = "Number of absences",
    y = "Final grade (G3)"
  )

```

Some correlations:

```{r}

num_vars <- dat %>%
  select(G1, G2, G3, age, absences)

cor(num_vars)
```

So correlation is very high between G1/G2/G3, this is expected.
Interestingly absences doesn't seem to have much correlation, and neither does age.